---
title: "Apatite REE - Draft code"
author: "Santiago Ramos Galarza"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# Load required libraries for data manipulation, visualization, and modeling

```{r load-packages, echo=TRUE, eval=T}
library(tidyverse)
library(tidymodels)
library(broom)
library(infer)
library(rpart)
library(rpart.plot)
```


# Loading data

```{r load-packages, echo=TRUE, eval=T}
setwd("/cloud/project/project-starter") 
df <- read_csv("data/apatite_data.csv")
```

# C1 Chondrite values (McDonough & Sun, 1995) [ppm]
```{r load-packages, echo=TRUE, eval=T}
chondrite <- c(La=0.237, Ce=0.612, Pr=0.0912, Nd=0.457, Sm=0.148, 
               Eu=0.0563, Gd=0.199, Tb=0.0361, Dy=0.246, Ho=0.0546, 
               Er=0.160, Tm=0.0247, Yb=0.161, Lu=0.0246)
```

# Normalizing REEs and calculate the Eu anomaly
```{r normalize, echo=TRUE, eval=T}
df_clean <- df %>%
  mutate(
    La_N = La139_ppm_mean / chondrite["La"],
    Ce_N = Ce140_ppm_mean / chondrite["Ce"],
    Pr_N = Pr141_ppm_mean / chondrite["Pr"],
    Nd_N = Nd146_ppm_mean / chondrite["Nd"],
    Sm_N = Sm147_ppm_mean / chondrite["Sm"],
    Eu_N = Eu153_ppm_mean / chondrite["Eu"],
    Gd_N = Gd157_ppm_mean / chondrite["Gd"],
    Eu_anomaly = Eu_N / sqrt(Sm_N * Gd_N),
    Tb_N = Tb159_ppm_mean / chondrite["Tb"],
    Dy_N = Dy163_ppm_mean / chondrite["Dy"],
    Ho_N = Ho165_ppm_mean / chondrite["Ho"],
    Er_N = Er166_ppm_mean / chondrite["Er"],
    Tm_N = Tm169_ppm_mean / chondrite["Tm"],
    Yb_N = Yb172_ppm_mean / chondrite["Yb"],
    Lu_N = Lu175_ppm_mean / chondrite["Lu"],
  )
glimpse(df_clean)
```


```{r summary, echo=TRUE, eval=T}
summary_table <- df_clean %>%
  group_by(Type) %>%
  summarize(
    Mean_Eu_Anomaly = mean(Eu_anomaly, na.rm = TRUE),
    SD_Eu_Anomaly = sd(Eu_anomaly, na.rm = TRUE),
    Count = n()
  )
summary_table
```
```{r box_plot, echo=TRUE, eval=T}
ggplot(df_clean, aes(x = Type, y = Eu_anomaly, fill = Type)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.shape = 1) +
  theme_linedraw() +
  labs(
    title = "Europium Anomaly Distribution: PCD vs. Barren Apatite",
    subtitle = "Higher values generally indicate higher oxygen fugacity",
    x = "Intrusive Rock Type",
    y = expression("Eu / Eu*")
  ) +
  scale_fill_manual(values = c("PCD" = "#4C72B0", "Barren" = "#C44E52")) +
  theme(legend.position = "none")
```

## Randomization Test

**Null Hypothesis ($H_0$):** There is no difference in the mean $Eu/Eu^*$ between PCD and Barren apatite. The variables are independent.
**Alternative Hypothesis ($H_A$):** There is a significant difference in the mean $Eu/Eu^*$ between the two groups.


```{r calculate-obs-stat, echo=TRUE, eval=TRUE}
obs_diff <- df_clean %>%
  specify(Eu_anomaly ~ Type) %>%
  calculate(stat = "diff in means", order = c("PCD", "Barren"))

obs_diff
```
# Null distribution

```{r null-dist, echo=TRUE, eval=TRUE}
set.seed(12)

null_dist <- df_clean %>%
  specify(Eu_anomaly ~ Type) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("PCD", "Barren"))

```


# Visualize
```{r visualize, echo=TRUE, eval=TRUE}
visualize(null_dist) +
  shade_p_value(obs_stat = obs_diff, direction = "two-sided") +
  theme_linedraw() +
  labs(
    title = "Simulation-Based Null Distribution",
    subtitle = "Red line: difference in means",
    x = "Simulated Difference in Means (PCD - Barren)",
    y = "Count"
  )
```
```{r p-value, echo=TRUE, eval=TRUE}
p_value <- null_dist %>%
  get_p_value(obs_stat = obs_diff, direction = "two-sided")

p_value
```

## 5. Estimating the Difference: Bootstrap Confidence Intervals

While the randomization test tells us the difference is significant, we want to quantify the magnitude of this difference. I will calculate a 95% Bootstrap Confidence Interval for the difference in means $(\bar{x}_{PCD} - \bar{x}_{Barren})$.

```{r bootstrap-ci, echo=TRUE, eval=TRUE}
set.seed(12)

# Bootstrap distribution
boot_dist <- df_clean %>%
  specify(Eu_anomaly ~ Type) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c("PCD", "Barren"))

# 95% CI
percentile_ci <- boot_dist %>%
  get_confidence_interval(level = 0.95, type = "percentile")

percentile_ci

# Visualize
visualize(boot_dist) +
  shade_confidence_interval(endpoints = percentile_ci) +
  theme_linedraw() +
  labs(
    title = "Bootstrap Distribution of Difference in Means",
    subtitle = "Shaded area represents the 95% Confidence Interval",
    x = "Difference in Mean Eu Anomaly (PCD - Barren)",
    y = "Count"
  )
```
## Simple Linear Model

slope as the difference between the PCD mean and the Barren mean

```{r linear-model, echo=TRUE, eval=TRUE}
eu_model <- lm(Eu_anomaly ~ Type, data = df_clean)
tidy(eu_model)
glance(eu_model)
```
```{r model-rmse, echo=TRUE, eval=TRUE}
model_results <- augment(eu_model)

rmse_calc <- model_results %>%
  summarize(RMSE = sqrt(mean(.resid^2)))

rmse_calc
```

## 7. Predictive Modeling: Classification Tree (Tidymodels)

This will evaluate all normalized REEs simultaneously to find the optimal thresholds for distinguishing fertile (PCD) from barren magmas.

```{r tidymodels-tree, echo=TRUE, eval=TRUE}

tree_data <- df_clean %>%
  select(Type, Eu_anomaly, La_N:Lu_N)

tree_spec <- decision_tree() %>%
  set_mode("classification") %>%
  set_engine("rpart")

tree_recipe <- recipe(Type ~ ., data = tree_data)

tree_workflow <- workflow() %>%
  add_recipe(tree_recipe) %>%
  add_model(tree_spec)


tree_fit <- tree_workflow %>%
  fit(data = tree_data)

# Plot
tree_engine <- extract_fit_engine(tree_fit)

rpart.plot(tree_engine, 
           main = "Classification Tree for Apatite Fertility",
           type = 3,           
           extra = 104,        
           box.palette = c("#C44E52", "#4C72B0"), 
           shadow.col = "gray",
           fallen.leaves = TRUE)
```
### Geochemical Overlap: REE Spider Diagram


```{r tidymodels-tree, echo=TRUE, eval=TRUE}
spider_data <- df_clean %>%
  mutate(Sample_ID = row_number()) %>% 
  select(Sample_ID, Type, ends_with("_N")) %>%
  pivot_longer(
    cols = ends_with("_N"), 
    names_to = "Element", 
    values_to = "Normalized_Value"
  ) %>%
  mutate(Element = str_remove(Element, "_N"))
```

```{r spide-diagram, echo=TRUE, eval=TRUE}
ree_order <- c("La", "Ce", "Pr", "Nd", "Sm", "Eu", "Gd", "Tb", "Dy", "Ho", "Er", "Tm", "Yb", "Lu")
spider_data$Element <- factor(spider_data$Element, levels = ree_order)

ggplot(spider_data, aes(x = Element, y = Normalized_Value, color = Type)) +
  geom_line(aes(group = Sample_ID), alpha = 0.15) +
  stat_summary(aes(group = Type), fun = mean, geom = "line", linewidth = 1.5) +
  stat_summary(aes(group = Type), fun = mean, geom = "point", size = 2) +
  scale_y_log10() + 
  theme_minimal() +
  labs(
    title = "Chondrite-Normalized REE Spider Diagram",
    y = "Sample / Chondrite",
    color = "Deposit Type"
  ) +
  scale_color_manual(values = c("PCD" = "#4C72B0", "Barren" = "#C44E52"))
```
## Comprehensive Analysis of All REEs

Instead of analyzing each Rare Earth Element individually, I will use functional programming (`purrr::map`) to iterate through all normalized REEs and the $Eu$ anomaly. This pipeline applies the simple linear model, the randomization test (1000 permutations), and the bootstrap confidence interval (1000 resamples) to every geochemical variable, compiling the results into a single summary table.

```{r analyze-all-ree, echo=TRUE, eval=TRUE}
# Group by the Element name
df_long <- df_clean %>%
  select(Type, Eu_anomaly, La_N:Lu_N) %>%
  pivot_longer(cols = -Type, names_to = "Element", values_to = "Value")

# Create a custom function that runs all tests on a single element
analyze_element <- function(sub_df) {
  set.seed(12)
  
  # Slope and R-Squared
  fit <- lm(Value ~ Type, data = sub_df)
  tidy_fit <- tidy(fit) %>% filter(term == "TypePCD")
  glance_fit <- glance(fit)
  
  # Observed Difference
  obs <- sub_df %>%
    specify(Value ~ Type) %>%
    calculate(stat = "diff in means", order = c("PCD", "Barren"))
  
  # Randomization Test
  null_dist <- sub_df %>%
    specify(Value ~ Type) %>%
    hypothesize(null = "independence") %>%
    generate(reps = 1000, type = "permute") %>%
    calculate(stat = "diff in means", order = c("PCD", "Barren"))
  
  p_val <- null_dist %>% get_p_value(obs_stat = obs, direction = "two-sided")
  
  # Bootstrap CI
  boot_dist <- sub_df %>%
    specify(Value ~ Type) %>%
    generate(reps = 1000, type = "bootstrap") %>%
    calculate(stat = "diff in means", order = c("PCD", "Barren"))
  
  ci <- boot_dist %>% get_confidence_interval(level = 0.95, type = "percentile")
  
  # Summarize
  tibble(
    Difference_in_Means = tidy_fit$estimate,
    R_Squared = glance_fit$r.squared,
    Rand_P_Value = p_val$p_value,
    CI_Lower = ci$lower_ci,
    CI_Upper = ci$upper_ci
  )
}

# Summary table
all_results <- df_long %>%
  nest(.by = Element) %>% # Group data by Element into 'lists'
  mutate(results = map(data, analyze_element)) %>% # Run the custom function
  unnest(results) %>% 
  select(-data) %>%
  arrange(Rand_P_Value, desc(R_Squared)) 

knitr::kable(all_results, digits = 4, 
             caption = "Statistical Summary for All Geochemical Proxies (PCD vs. Barren)")
```
